---
title: "Plotting MEFEC temperature data"
author: "K. Zamzow"
date: "`r Sys.Date()`"
output: html_document
---
#This scrip places all the plots in folders for easy viewing
```{r setup, include=FALSE}

knitr::opts_chunk$set(echo = FALSE, message =FALSE, warning=FALSE, error = FALSE, eval=FALSE)
```

#install packages and directories
```{r pac.ages, eval=FALSE}
install.packages(c("httr", "jsonlite", "dplyr"))
```

```{r libraries}

library(httr)
library(jsonlite)
library(dplyr)
```

```{r directories}
base_dir<-"~/Chemical & physical environments/Alaska weather/Mat Valley weather"
```

#retrieve data directly from xmacis (need internet)
```{r set up to get data directly from xmacis}
#This is pulled from the website and will only work if internet is on
#from AI
get_xmacis_data <- function(sid, sdate, edate, elems) {
  url <- "http://data.rcc-acis.org/StnData"
  
  body <- list(
    sid = sid,
    sdate = sdate,
    edate = edate,
    elems = elems
  )
  
  response <- POST(url, body = body, encode = "json")
  content <- content(response, "text")
  data <- fromJSON(content)
  
  return(data$data)
}

```

```{r run function to pull data in}
#from AI
# Set parameters
station_id <- "505733"  # This should be MEFEC
start_date <- "1917-07-05"
end_date <- "2025-01-31"
elements <- list(list(name="maxt"), list(name="mint"))

# Fetch data
temperature_data<- get_xmacis_data(station_id, start_date, end_date, elements)

#convert to data frame
temperature_df<-as.data.frame(temperature_data)

colnames(temperature_df) <- c("date", "max_temp", "min_temp")

# Display first few rows
head(temperature_df)
```

```{r save the data}
#from AI

# Save as CSV in the base_dir
write.csv(temperature_df, "MEFEC2025.02.15Temps.csv", row.names = FALSE)

```

```{r ensure temperatures are seen as numbers}

Tmin<-as.numeric(temperature_df$min_temp, na.rm=TRUE)
Tmax<-as.numeric(temperature_df$max_temp, na.rm=TRUE)
```

#Saving plots in folders
```{r How to save plots, eval=FALSE}
#create directories
dir.create("annual_plots_temp", showWarnings = FALSE)
dir.create("monthly_plots_tmax", showWarnings = FALSE)
dir.create("monthly_plots_tmin", showWarnings = FALSE)

#create lists to store them in
annual_plots_temp <- list()
monthly_plots_tmax <- list()
monthly_plots_tmin <- list()

#modify the ggsave
  ggsave(
    filename = paste0("monthly_plots_tmax/", month, "_tmax_plot.png"),
    plot = plot_tmax,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
  
# Store Tmax plot in the list
  monthly_plots_tmax[[month]] <- plot_tmax

# Repeat the process for Tmin data
  MonthTmin <- MonthTemp %>%
    dplyr::select(date, min_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
```

#Plot annual 30-yr rolling averages
```{r create a Month column from date, results='hide'}

library (lubridate)
library (dplyr)

temperature_df<-temperature_df%>%
  mutate(
    Year = year(temperature_df$date),
    Month = month(temperature_df$date, label = TRUE, abbr = TRUE)
  )
#abbr=TRUE puts all the dates as 3-letter abbreviations

```

```{r install package for rolling 30 year average, eval=FALSE}
#from AI
install.packages(tibbletime)
```

```{r annual TMin rolling 30 year average}

library(tibbletime)
library(zoo)

#set up storage
dir.create("annual_plots_temp", showWarnings = FALSE)
annual_plots_temp <- list()

# Prepare the data
Tmin2<- temperature_df%>%
  dplyr::select(date, min_temp) %>%
  mutate(
    Year = year(date),
    Month = month(date, label = TRUE, abbr = TRUE)
  )

#chec/ for NAs, remove them, and count how many were removed
Tmin2 <- Tmin2 %>%
  filter(!is.na(min_temp))
Tmin2$min_temp <- as.numeric(Tmin2$min_temp)
sum(is.na(Tmin2$min_temp))

# Function to calculate 30-year average
calculate_30yr_avg <- function(data, start_year) {
  data %>%
    filter(Year >= start_year, Year < start_year + 30) %>%
    summarise(min_temp_30yr = mean(min_temp, na.rm = TRUE))
}

# Calculate for each decade
decades <- seq(1917, max(Tmin2$Year) - 29, by = 10)

results<- data.frame()

for (decade in decades) {
  avg <- calculate_30yr_avg(Tmin2, decade)
  results<- rbind(results, data.frame(start_year = decade, end_year = decade + 29, min_temp_30yr = avg$min_temp_30yr))
}

# View results
print(results)
```

```{r plot annual TMin}
library(ggplot2)

#this is saved "manually" by exporting the image into folder "annual_plots_temp" instead of trying to automatically code it into a folder, which was having problems.

ggplot(data=results)+aes(x=start_year, y=min_temp_30yr)+
  geom_point()+
  geom_smooth(method="lm", se=FALSE, color="blue")+
  ylim(25,30)+
  xlab("")+ylab("Temperature (F)")+
  ggtitle("Minimum annual average temperatures 1917-2024", subtitle= "as 30-year averages by decade")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))+
  theme(axis.text=element_text(size=12))+
  theme(title = element_text(size=14))
```

```{r save plots in folder, eval=FALSE}
#use ggsave to create a file for storage
#this doesn't seem to wor/ for annual, need to figure it out
  ggsave(
    filename = paste0("annual_plots_temp/", month, "_temp_plot.png"),
    plot = plot_temp,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
  
# Store Tmax plot in the list
  annual_plots_temp[[Year]] <- plot_temp
```

```{r annual TMax rolling 30 year average}

#this gets to a list of results but no plot
# Prepare the data
Tmax2 <- temperature_df %>%
  dplyr::select(date, max_temp) %>%
  mutate(
    Year = year(date),
    Month = month(date, label = TRUE, abbr = TRUE)
  )

Tmax2$max_temp <- as.numeric(Tmax2$max_temp)

# Function to calculate 30-year average
calculate_30yr_avg <- function(data, start_year) {
  data %>%
    filter(Year >= start_year, Year < start_year + 30) %>%
    summarise(max_temp_30yr = mean(max_temp, na.rm = TRUE))
}

# Calculate for each decade
decades <- seq(1917, max(Tmax2$Year) - 29, by = 10)

resultsTmax<- data.frame()
for (decade in decades) {
  avg <- calculate_30yr_avg(Tmax2, decade)
  resultsTmax <- rbind(resultsTmax, data.frame(start_year = decade, end_year = decade + 29, max_temp_30yr = avg$max_temp_30yr))
}

# View results
print(resultsTmax)
```

```{r plot annual TMax}
library(ggplot2)

#manually exported image into annual_temps folder
ggplot(data=resultsTmax)+aes(x=start_year, y=max_temp_30yr)+
  geom_point()+
  geom_smooth(method="lm", se=FALSE, color="blue")+
  ylim(41, 46)+
  xlab("")+ylab("Temperature (F)")+
  ggtitle("Maximum annual average temperatures 1917-2024", subtitle= "as 30-year averages by decade")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))+
  theme(axis.text=element_text(size=12))+
  theme(title = element_text(size=14))

```

#Plot 30-yr rolling averages for an individual month
```{r select a month for analysis}

DecTemp<-temperature_df%>%
  filter(Month == "Dec")
DecTemp <- DecTemp %>%
  filter(!is.na(max_temp))
DecTemp$max_temp <- as.numeric(DecTemp$max_temp)

write.csv(temperature_df, "MEFEC2025.02.15DecTemps.csv", row.names = FALSE)
```

```{r Dec max temps}
#this gets to a list of results but no plot
# Prepare the data
DecTmax <- DecTemp%>%
  dplyr::select(date, max_temp) %>%
  mutate(
    Year = year(date),
    Month = month(date, label = TRUE, abbr = TRUE)
  )

# Function to calculate 30-year average
calculate_30yr_avg <- function(data, start_year) {
  data %>%
    filter(Year >= start_year, Year < start_year + 30) %>%
    summarise(max_temp_30yr = mean(max_temp, na.rm = TRUE))
}

# Calculate for each decade
decades <- seq(1917, max(DecTmax$Year) - 29, by = 10)

resultsDecTmax<- data.frame()
for (decade in decades) {
  avg <- calculate_30yr_avg(DecTmax, decade)
  resultsDecTmax <- rbind(resultsDecTmax, data.frame(start_year = decade, end_year = decade + 29, max_temp_30yr = avg$max_temp_30yr))
}

# View results
print(resultsDecTmax)
```

```{r plot DecTmax}
library(ggplot2)
#manually exported this into Tmax folder

ggplot(data=resultsDecTmax)+aes(x=start_year, y=max_temp_30yr)+
  geom_point()+  
  geom_smooth(method="lm", se=FALSE, color="blue")+
  ylim(10, 30)+
  xlab("")+ylab("Temperature (F)")+
  ggtitle("Maximum December average temperatures 1917-2024", subtitle= "as 30-year averages by decade")+
  theme_bw()+
  theme(axis.text.x=element_text(angle=90, hjust=1, vjust=0.5))+
  theme(axis.text=element_text(size=12))+
  theme(title = element_text(size=14))

```

```{r monthly Tmax and Tmin from AI placed in folders, eval=FALSE}
#this goes all the way to placing monthly plots in Tmin and Tmax folders. It should wor/ but I created two separate chun/s instead

#you need to specify months again
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")
print(months)

#and ensure values are seen as numbers again
temperature_df$max_temp <- as.numeric(temperature_df$max_temp)
temperature_df$min_temp <- as.numeric(temperature_df$min_temp)


#create directories
dir.create("monthly_plots_tmax", showWarnings = FALSE)
dir.create("monthly_plots_tmin", showWarnings = FALSE)

# Create empty lists to store plots
monthly_plots_tmax <- list()
monthly_plots_tmin <- list()

# Start the for-loop
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare and plot Tmax data
  MonthTmax <- MonthTemp %>%
    dplyr::select(date, max_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmax$max_temp <- as.numeric(MonthTmax$max_temp)
  
  # Calculate 30-year average for Tmax
  decades <- seq(1917, max(MonthTmax$Year) - 29, by = 10)
  resultsMonthTmax <- data.frame()
  
  for (decade in decades) {
    avg <- MonthTmax %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(max_temp_30yr = mean(max_temp, na.rm = TRUE))
    
    resultsMonthTmax <- rbind(resultsMonthTmax, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         max_temp_30yr = avg$max_temp_30yr))
  }
  
  # Calculate y-axis limits for Tmax
  y_min_max <- floor(min(resultsMonthTmax$max_temp_30yr))
  y_max_max <- y_min_max + 5
  
  # Create Tmax plot
  plot_tmax <- ggplot(data = resultsMonthTmax) +
    aes(x = start_year, y = max_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    ylim(y_min_max, y_max_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Maximum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Save Tmax plot
  ggsave(
    filename = paste0("monthly_plots_tmax/", month, "_tmax_plot.png"),
    plot = plot_tmax,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
  
  # Store Tmax plot in the list
  monthly_plots_tmax[[month]] <- plot_tmax
  
  # Repeat the process for Tmin data
  MonthTmin <- MonthTemp %>%
    dplyr::select(date, min_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmin$min_temp <- as.numeric(MonthTmin$min_temp)
  
  # Calculate 30-year average for Tmin
  resultsMonthTmin <- data.frame()
  
  for (decade in decades) {
    avg <- MonthTmin %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(min_temp_30yr = mean(min_temp, na.rm = TRUE))
    
    resultsMonthTmin <- rbind(resultsMonthTmin, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         min_temp_30yr = avg$min_temp_30yr))
  }
  
  # Calculate y-axis limits for Tmin
  y_min_min <- floor(min(resultsMonthTmin$min_temp_30yr))
  y_max_min <- y_min_min + 5
  
  # Create Tmin plot
  plot_tmin <- ggplot(data = resultsMonthTmin) +
    aes(x = start_year, y = min_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ylim(y_min_min, y_max_min) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Minimum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Save Tmin plot
  ggsave(
    filename = paste0("monthly_plots_tmin/", month, "_tmin_plot.png"),
    plot = plot_tmin,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
  
  # Store Tmin plot in the list
  monthly_plots_tmin[[month]] <- plot_tmin
}

```

```{r AI for monthly Tmax plots to folders}
#this wor.s
# Create a vector of all months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Create an empty list to store Tmax plots
monthly_plots_tmax <- list()

# Create directory for Tmax plots
dir.create("monthly_plots_tmax", showWarnings = FALSE)

#ensure values are seen as numeric
temperature_df$max_temp<-as.numeric(temperature_df$max_temp)

# Start the for-loop for Tmax
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare the data for max temperatures
  MonthTmax <- MonthTemp %>%
    dplyr::select(date, max_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmax$max_temp <- as.numeric(MonthTmax$max_temp)
  
  # Calculate 30-year average
  decades <- seq(1917, max(MonthTmax$Year) - 29, by = 10)
  resultsMonthTmax <- data.frame()
  
  for (decade in decades) {
    avg <- MonthTmax %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(max_temp_30yr = mean(max_temp, na.rm = TRUE))
    
    resultsMonthTmax <- rbind(resultsMonthTmax, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         max_temp_30yr = avg$max_temp_30yr))
  }
  
    
  # Calculate y-axis limits
  y_min <- floor(min(resultsMonthTmax$max_temp_30yr))
  y_max <- y_min + 5
  
  # Create plot for the current month
  plot <- ggplot(data = resultsMonthTmax) +
    aes(x = start_year, y = max_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "red") +
    ylim(y_min, y_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Maximum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Save the plot
  ggsave(
    filename = paste0("monthly_plots_tmax/", month, "_tmax_plot.png"),
    plot = plot,
    width = 10,
    height = 6,
    units = "in",
  )
  
  # Store the plot in the list
  monthly_plots_tmax[[month]] <- plot
}
```

```{r AI for monthly Tmin plots to folders}
# This runs but is not placing plots in the monthly_plots_tmin folder. It had no problem putting them in a monthly_tmin_plots folder though??

# Create a vector of all months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

#Create an empty list to store Tmin plots
monthly_plots_tmin <- list()

# Create directory for Tmin plots
dir.create("monthly_plots_tmin", showWarnings = FALSE)

#ma.e sure values are seen as numbers
temperature_df$min_temp<-as.numeric(temperature_df$min_temp)

# Start the for-loop for Tmin
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare the data for min temperatures
  MonthTmin <- MonthTemp %>%
    dplyr::select(date, min_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmin$min_temp <- as.numeric(MonthTmin$min_temp)
  
  # Calculate 30-year average
  decades <- seq(1917, max(MonthTmin$Year) - 29, by = 10)
  resultsMonthTmin <- data.frame()
  
  for (decade in decades) {
    avg <- MonthTmin %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(min_temp_30yr = mean(min_temp, na.rm = TRUE))
    
    resultsMonthTmin <- rbind(resultsMonthTmin, 
                              data.frame(start_year = decade,
                                         end_year = decade + 29,
                                         min_temp_30yr =
                                           avg$min_temp_30yr))
  }
  
  # Calculate y-axis limits
  y_min <- floor(min(resultsMonthTmin$min_temp_30yr))
  y_max <- y_min + 5
  
  # Create plot for the current month
  plot <- ggplot(data = resultsMonthTmin) +
    aes(x = start_year, y = min_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ylim(y_min, y_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Minimum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Save the plot
  ggsave(
    filename = paste0("monthly_plots_tmin/", month, "_tmin_plot.png"),
    plot = plot,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
  
  # Store the plot in the list
  monthly_plots_tmin[[month]] <- plot
}

```

#Plot 30-yr rolling averages for all months, view as grid
```{r Tmax for loop for each month gridplot}
#this creates a grid of each individual month with Tmax rolling average. Store it by manually exporting to the Tmax folder. Change dimensions to be 1600 wide instead of 600 wide.

# First, let's create a vector of all months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Create an empty list to store plots
monthly_plots <- list()

# Start the for-loop
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare the data for max temperatures
  MonthTmax <- MonthTemp %>%
    dplyr::select(date, max_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmax$max_temp <- as.numeric(MonthTmax$max_temp)
  
  # Calculate 30-year average
  decades <- seq(1917, max(MonthTmax$Year) - 29, by = 10)
  resultsMonthTmax <- data.frame()
  
  for (decade in decades) {
    avg <- MonthTmax %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(max_temp_30yr = mean(max_temp, na.rm = TRUE))
    
    resultsMonthTmax <- rbind(resultsMonthTmax, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         max_temp_30yr = avg$max_temp_30yr))
  }
  # Calculate y-axis limits
  y_min <- floor(min(resultsMonthTmax$max_temp_30yr))
  y_max <- y_min + 5
  
  # Create plot for the current month
  plot <- ggplot(data = resultsMonthTmax) +
    aes(x = start_year, y = max_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ylim(y_min, y_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Maximum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Store the plot in the list
  monthly_plots[[month]] <- plot
}

# Display all plots (you might want to adjust the layout for better visibility)
library(gridExtra)
do.call(grid.arrange, c(monthly_plots, ncol = 3))

```

```{r Tmin for loop for each month gridplot}

# First, let's create a vector of all months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Create an empty list to store plots
monthly_plots <- list()

# Start the for-loop
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare the data for max temperatures
  MonthTmin <- MonthTemp %>%
    dplyr::select(date, min_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmin$min_temp <- as.numeric(MonthTmin$min_temp)
  
  # Calculate 30-year average
  decades <- seq(1917, max(MonthTmin$Year) - 29, by = 10)
  resultsMonthTmin <- data.frame()
  
  for (decade in decades) {
    avg <- MonthTmin %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(min_temp_30yr = mean(min_temp, na.rm = TRUE))
    
    resultsMonthTmin <- rbind(resultsMonthTmin, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         min_temp_30yr = avg$min_temp_30yr))
  }
  # Calculate y-axis limits
  y_min <- floor(min(resultsMonthTmin$min_temp_30yr))
  y_max <- y_min + 5
  
  # Create plot for the current month
  plot <- ggplot(data = resultsMonthTmin) +
    aes(x = start_year, y = min_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ylim(y_min, y_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Minimum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Store the plot in the list
  monthly_plots[[month]] <- plot
}

# Display all plots (you might want to adjust the layout for better visibility)
library(gridExtra)
do.call(grid.arrange, c(monthly_plots, ncol = 3))

```


#Plot 30-yr rolling averages for all months, view individually

```{r run code to save each Tmin month plot individually, eval=FALSE}

#superseded by line 321 code but /eep for eventually ta/ing out line 321 chun/ and modifying these chun/s
# First, let's create a vector of all months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Create an empty list to store plots
Tminmonthly_plots <- list()

#create a directory to store them in
dir.create("monthly_plots", showWarnings = FALSE)

# Start the for-loop
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare the data for max temperatures
  MonthTmin <- MonthTemp %>%
    dplyr::select(date, min_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmin$min_temp <- as.numeric(MonthTmin$min_temp)
  
# Calculate 30-year average
  decades <- seq(1917, max(MonthTmin$Year) - 29, by = 10)
  
  resultsMonthTmin <- data.frame()

  for (decade in decades) {
    avg <- MonthTmin %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(min_temp_30yr = mean(min_temp, na.rm = TRUE))
    
    resultsMonthTmin <- rbind(resultsMonthTmin, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         min_temp_30yr = avg$min_temp_30yr))
  }
  
# Calculate y-axis limits
  y_min <- floor(min(resultsMonthTmin$min_temp_30yr))
  y_max <- y_min + 5
  
# Create plot for the current month
  plot <- ggplot(data = resultsMonthTmin) +
    aes(x = start_year, y = min_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ylim(y_min, y_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("Minimum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Save the plot
  ggsave(
    filename = paste0("monthly_plots/", month, "_temperature_plot.png"),
    plot = plot,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
    # Store the plot in the list (if you still want to display them all together)
  Tminmonthly_plots[[month]] <- plot
}

print(Tminmonthly_plots)
```

```{r run code to save each Tmax month plot individually, eval=FALSE}
#superseded by line 321 code but /eep for eventually removing that chun/ and modifying this one
# First, let's create a vector of all months
months <- c("Jan", "Feb", "Mar", "Apr", "May", "Jun", "Jul", "Aug", "Sep", "Oct", "Nov", "Dec")

# Create an empty list to store plots
Tmaxmonthly_plots <- list()

#create a directory to store them in
dir.create("monthly_plots", showWarnings = FALSE)

# Start the for-loop
for (month in months) {
  # Filter data for the current month
  MonthTemp <- temperature_df %>%
    filter(Month == month)
  
  # Prepare the data for max temperatures
  MonthTmax <- MonthTemp %>%
    dplyr::select(date, max_temp) %>%
    mutate(
      Year = year(date),
      Month = month(date, label = TRUE, abbr = TRUE)
    )
  
  MonthTmax$max_temp <- as.numeric(MonthTmax$max_temp)
  
# Calculate 30-year average
  decades <- seq(1917, max(MonthTmax$Year) - 29, by = 10)
  
  resultsMonthTmax <- data.frame()

  for (decade in decades) {
    avg <- MonthTmax %>%
      filter(Year >= decade, Year < decade + 30) %>%
      summarise(max_temp_30yr = mean(max_temp, na.rm = TRUE))
    
    resultsMonthTmax <- rbind(resultsMonthTmax, 
                              data.frame(start_year = decade, 
                                         end_year = decade + 29, 
                                         max_temp_30yr = avg$max_temp_30yr))
  }
  
# Calculate y-axis limits
  y_min <- floor(min(resultsMonthTmax$max_temp_30yr))
  y_max <- y_min + 5
  
# Create plot for the current month
  plot <- ggplot(data = resultsMonthTmax) +
    aes(x = start_year, y = max_temp_30yr) +
    geom_point() +
    geom_smooth(method = "lm", se = FALSE, color = "blue") +
    ylim(y_min, y_max) +
    xlab("") + ylab("Temperature (F)") +
    ggtitle(paste("maximum", month, "average temperatures 1917-2024"),
            subtitle = "as 30-year averages by decade") +
    theme_bw() +
    theme(axis.text.x = element_text(angle = 90, hjust = 1, vjust = 0.5),
          axis.text = element_text(size = 12),
          title = element_text(size = 14))
  
  # Save the plot
  ggsave(
    filename = paste0("monthly_plots/", month, "_temperature_plot.png"),
    plot = plot,
    width = 10,
    height = 6,
    units = "in",
    dpi = 300
  )
    # Store the plot in the list (if you still want to display them all together)
  Tmaxmonthly_plots[[month]] <- plot
}

print(Tmaxmonthly_plots)
```

```{r set up to save each month plot individually, eval=FALSE}
#ditto - superseded by the AI version on line 321, but /eep this to brea/ out Tmax monthly, Tmin monthly, and the annual temps eventually 

#create a directory to store them in
dir.create("monthly_plots", showWarnings = FALSE)

# Inside the for-loop, after creating the plot
ggsave(
  filename = paste0("monthly_plots/", month, "_temperature_plot.png"),
  plot = plot,
  width = 10,
  height = 6,
  units = "in",
  dpi = 300
)
```

